DOpElib:\\ Differential Equations and Optimization Environment

### 2013-12-12 Examples/OPT/StatPDE/Example1/content.tex
cosmetic changes in examples

Moreover, we would like to work in the Hilbert space $Q$. However, the derivative 
$j'(q) \in H^*$ only. Hence, we need to calculate the Riesz representation 
for the gradient $\nabla j(q) \in H$ using the definition:
\[
 (\nabla j(q),\delta q)_Q = j'(q)(\delta q) \quad \forall\,\delta q\in Q.
\]
In the given example, $Q = L^2(\Omega)$ and hence the scalar product will be the
standard $L^2$-inner product. Simillarly, we can define the Hessian operator 
$H(q) \in mathcal L(Q,Q)$ by defining
\[
 (H(q)\tau q,\delta q)_Q = j''(q)(\tau q,\delta q) \quad \forall\,\delta q,\tau q\in Q.
\]
}

\subsubsection{Implementation in \dope{}}
{
From the previous details, and the definition of the Lagrangian and its 
derivatives it is clear, that the user has to provide the respective derivatives. 
Since the Lagrangian consists of the PDE and the cost functional it 
is sufficient to provide the respective derivatives, while \texttt{DOpElib} 
will assemble them as required.
Test functions for vector valued terms will be denoted by an index $i$ while 
matrix valued terms are indexed in $i$ and $j$. Test functions in the control 
space $Q$ are denoted as $\phi^q_i$ while those in the state space $V$ are denoted 
as $\phi_i$.

To solve the linear equations the following matrices are needed
\begin{align*}
\text{ElementMatrix} \quad 
&\Leftrightarrow \quad a_{i,j} = a'_u(q,u)(\phi_j,\phi_i), \\
\text{ControlElementMatrix } \quad 
&\Leftrightarrow \quad a_{i,j} = (\phi^q_j, \phi^q_i)_Q.
\end{align*}
The first one is required for all primal and dual PDE solves, while the second one is needed 
to calculate the Riesz representation of the derivatives of $j$. 
If desired, the matrix for the adjoint PDEs can be provided separately as 
\texttt{ElementMatrix\underline{ }T}, but otherwise this will be 
calculated automatically from the primal matrix.

Additional terms are needed to calculate the corresponding 
right hand sides. These are for the PDE
\begin{align*}
\text{ElementEquation (state)} \quad &\Leftrightarrow \quad a(q,u)(\phi_i), \\
\text{ElementEquation\underline{ }U (adjoint)} \quad 
&\Leftrightarrow \quad a_u'(q,u)(\phi_i, z), \\ 
\text{ElementEquation\underline{ }UT (tangent)} \quad 
&\Leftrightarrow \quad a_u'(q,u)(\delta u, \phi_i), \\
\text{ElementEquation\underline{ }UTT (adjoint hessian)} \quad 
&\Leftrightarrow \quad  a_u'(q,u)(\phi_i, \delta z), \\
\text{ElementEquation\underline{ }Q (gradient)} \quad 
&\Leftrightarrow \quad a_q'(q,u)(\phi^q_i,z), \\
\text{ElementEquation\underline{ }QT (tangent)} \quad 
&\Leftrightarrow \quad a_q'(q,u)(\delta q, \phi_i), \\
\text{ElementEquation\underline{ }QTT (hessian)} \quad 
&\Leftrightarrow \quad a_{q}'(q,u)(\phi^q_i,\delta z), \\
\text{ElementEquation\underline{ }UU (adjoint hessian)} \quad 
&\Leftrightarrow \quad a_{uu}''(q,u)(\delta u, \phi_i, z), \\
\text{ElementEquation\underline{ }QU (adjoint hessian)} \quad 
&\Leftrightarrow \quad a_{qu}''(q,u)(\delta q, \phi_i, z), \\
\text{ElementEquation\underline{ }UQ (hessian)} \quad 
&\Leftrightarrow \quad a_{uq}''(q,u)(\delta u, \phi^q_i, z),\\
\text{ElementEquation\underline{ }QQ (hessian)} \quad 
&\Leftrightarrow \quad a_{qq}''(q,u)(\delta q, \phi^q_i, z),\\
\text{ElementRightHandSide (state)} \quad 
&\Leftrightarrow \quad f(\phi_i), \\
\text{ControlElementEquation (gradient or hessian)} \quad 
&\Leftrightarrow \quad (\nabla j(q),\phi^q_i)_Q.
\end{align*}
As for PDE problems, it is up to the user to decide if the ElementRightHandSide
is used, or if the terms are included in the ElementEquation.

For the cost functional, we have to provide
\begin{align*}
\text{ElementValue (all)} \quad  & \Leftrightarrow \quad J(q,u), \\
\text{ElementValue\underline{ }U (all)} \quad  & \Leftrightarrow \quad J_u'(q,u)(\phi_i), \\
\text{ElementValue\underline{ }Q (all)} \quad  & \Leftrightarrow \quad J_q'(q,u)(\phi_i), \\
\text{ElementValue\underline{ }UU (all)} \quad & \Leftrightarrow \quad J_{uu}''(q,u)(\delta u, \phi_i), \\
\text{ElementValue\underline{ }QU (all)} \quad & \Leftrightarrow \quad J_{qu}''(q,u)(\delta q, \phi_i), \\
\text{ElementValue\underline{ }UQ (all)} \quad & \Leftrightarrow \quad J_{uq}''(q,u)(\delta u, \phi^q_i), \\
\text{ElementValue\underline{ }QQ (all)} \quad & \Leftrightarrow \quad J_{qq}''(q,u)(\delta q, \phi^q_i).
\end{align*}

@@

\subsubsection{Back to the specific equations in this example}
We have
\begin{align*}
a(q,u)(\phi) &= (\nabla u, \nabla\phi) - (q + f, \phi),\\
a_u'(q,u)(\phi, z) &= (\nabla\phi, \nabla z),\\
a_u'(q,u)(\delta u, \phi) &= (\nabla\delta u, \nabla\phi),\\
a_u'(q,u)(\phi, \delta z) &= (\nabla\phi, \nabla\delta z),\\
a_q'(q,u)(\delta q, \phi) &= -(z,\psi^q),\\
a_q'(q,u)(\delta q, \phi) &= (\delta q,\phi),\\
a_{q}'(q,u)(\delta q,\delta z) &= - (\delta z,\psi^q).
\end{align*}
%
For the cost functional, we have the following terms:
\begin{align*}
J(q,u) &= \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2,\\
J_u'(q,u)(\phi) = (u-u^d, \phi),\\
J_q'(q,u)(\phi) = \alpha (q,\psi^q),\\
J_{uu}''(q,u)(\psi, \phi) = (\delta u, \phi).
\end{align*}
All other terms, specifically mixed terms with $QU$ etc. are zero in this
example.

@@

}

>>>>

Moreover, we would like to work in the Hilbert space $Q$. However, the derivative 
$j'(q) \in H^*$ only. Hence, we need to calculate the Riesz representation 
for the gradient $\nabla j(q) \in H$ using the definition:
\[
 (\nabla j(q),\delta q)_Q = j'(q)(\delta q) \quad \forall\,\delta q\in Q.
\]
In the given example, $Q = L^2(\Omega)$ and hence the scalar product will be the
standard $L^2$-inner product. Simillarly, we can define the Hessian operator 
$H(q) \in \mathcal L(Q,Q)$ by defining
\[
 (H(q)\tau q,\delta q)_Q = j''(q)(\tau q,\delta q) \quad \forall\,\delta q,\tau q\in Q.
\]
}

\subsubsection{Implementation in \dope{}}
From the previous details, and the definition of the Lagrangian and its 
derivatives it is clear, that the user has to provide the respective derivatives. 
Since the Lagrangian consists of the PDE and the cost functional it 
is sufficient to provide the respective derivatives, while \texttt{DOpElib} 
will assemble them as required.
Test functions for vector valued terms will be denoted by an index $i$ while 
matrix valued terms are indexed in $i$ and $j$. Test functions in the control 
space $Q$ are denoted as $\phi^q_i$ while those in the state space $V$ are denoted 
as $\phi_i$.

To solve the linear equations the following matrices are needed{\color{red}
\begin{align*}
&\texttt{ElementMatrix} 
&\Leftrightarrow \quad a_{i,j} &= a'_u(q,u)(\phi_j,\phi_i), \\
&\texttt{ControlElementMatrix}  
&\Leftrightarrow \quad a_{i,j} &= (\phi^q_j, \phi^q_i)_Q.
\end{align*}}
The first one is required for all primal and dual PDE solves, while the second one is needed 
to calculate the Riesz representation of the derivatives of $j$. 
If desired, the matrix for the adjoint PDEs can be provided separately as 
\texttt{ElementMatrix\underline{ }T}, but otherwise this will be 
calculated automatically from the primal matrix.

Additional terms are needed to calculate the corresponding 
right hand sides. These are for the PDE the following: 
{\color{red}
\begin{align*}
&\texttt{ElementEquation}\text{ (state)} &\Leftrightarrow &\quad a(q,u)(\phi_i), \\
&\texttt{ElementRightHandSide}\text{ (state)}  &\Leftrightarrow &\quad f(\phi_i), \\
&\texttt{ControlElementEquation}\text{ (gradient or hessian)}  &\Leftrightarrow &\quad (\nabla j(q),\phi^q_i)_Q,
\end{align*}
as well as 
\begin{align*}
&\texttt{ElementEquation\underline{ }U}\text{ (adjoint)} 
&\Leftrightarrow &\quad a_u'(q,u)(\phi_i, z),\\
&\texttt{ElementEquation\underline{ }Q}\text{ (gradient)} 
&\Leftrightarrow &\quad a_q'(q,u)(\phi^q_i,z), \\
\end{align*}
the terms
\begin{align*}
&\texttt{ElementEquation\underline{ }UU}\text{ (adjoint hessian)}  
&\Leftrightarrow &\quad a_{uu}''(q,u)(\delta u, \phi_i, z),\\
&\texttt{ElementEquation\underline{ }UQ}\text{ (hessian)}
&\Leftrightarrow &\quad a_{uq}''(q,u)(\delta u, \phi^q_i, z),\\
&\texttt{ElementEquation\underline{ }QU}\text{ (adjoint hessian)} 
&\Leftrightarrow &\quad a_{qu}''(q,u)(\delta q, \phi_i, z), \\
&\texttt{ElementEquation\underline{ }QQ}\text{ (hessian)} 
&\Leftrightarrow &\quad a_{qq}''(q,u)(\delta q, \phi^q_i, z),\\
&\texttt{ElementEquation\underline{ }UT}\text{ (tangent)}  
&\Leftrightarrow &\quad a_u'(q,u)(\delta u, \phi_i), \\
&\texttt{ElementEquation\underline{ }QT}\text{ (tangent)}  
&\Leftrightarrow &\quad a_q'(q,u)(\delta q, \phi_i),
\end{align*}
and finally
\begin{align*}
&\texttt{ElementEquation\underline{ }UTT}\text{ (adjoint hessian)} \quad 
&\Leftrightarrow &\quad  a_u'(q,u)(\phi_i, \delta z), \\
&\texttt{ElementEquation\underline{ }QTT}\text{ (hessian)} \quad 
&\Leftrightarrow &\quad a_{q}'(q,u)(\phi^q_i,\delta z).
\end{align*}}
As for PDE problems, it is up to the user to decide if the \texttt{ElementRightHandSide}
is used, or if the terms are included in the \texttt{ElementEquation}.

For the cost functional, we have to provide{\color{red}
\begin{align*}
&\texttt{ElementValue}\text{ (all)}   &\Leftrightarrow &\quad J(q,u), \\
&\texttt{ElementValue\underline{ }U}\text{ (all)}    &\Leftrightarrow &\quad J_u'(q,u)(\phi_i), \\
&\texttt{ElementValue\underline{ }Q}\text{ (all)}    &\Leftrightarrow &\quad J_q'(q,u)(\phi_i), \\
&\texttt{ElementValue\underline{ }UU}\text{ (all)}  &\Leftrightarrow &\quad J_{uu}''(q,u)(\delta u, \phi_i), \\
&\texttt{ElementValue\underline{ }UQ}\text{ (all)}  &\Leftrightarrow &\quad J_{uq}''(q,u)(\delta u, \phi^q_i), \\
&\texttt{ElementValue\underline{ }QU}\text{ (all)}   &\Leftrightarrow &\quad J_{qu}''(q,u)(\delta q, \phi_i), \\
&\texttt{ElementValue\underline{ }QQ}\text{ (all)}  &\Leftrightarrow &\quad J_{qq}''(q,u)(\delta q, \phi^q_i).
\end{align*}}

@@

\subsubsection{Back to the specific equations in this example}
We have
\begin{align*}
a(q,u)(\phi) &= (\nabla u, \nabla\phi) - (q + f, \phi),\\
a_u'(q,u)(\phi, z) &= (\nabla\phi, \nabla z),\\
a_u'(q,u)(\delta u, \phi) &= (\nabla\delta u, \nabla\phi),\\
a_u'(q,u)(\phi, \delta z) &= (\nabla\phi, \nabla\delta z),\\
a_q'(q,u)(\delta q, \phi) &= -(z,\psi^q),\\
a_q'(q,u)(\delta q, \phi) &= (\delta q,\phi),\\
a_{q}'(q,u)(\delta q,\delta z) &= - (\delta z,\psi^q).
\end{align*}
%
For the cost functional, we have the following terms:
\begin{align*}
J(q,u) &= \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2,\\
J_u'(q,u)(\phi) &= (u-u^d, \phi),\\
J_q'(q,u)(\phi) &= \alpha (q,\psi^q),\\
J_{uu}''(q,u)(\psi, \phi) &= (\delta u, \phi).
\end{align*}
All other terms, specifically mixed terms with $QU$ etc. are zero in this
example.

@@

### 2013-12-12 Examples/OPT/StatPDE/Example2/content.tex
cosmetic changes in examples

\subsubsection{General problem description}
This example solves the following pointwise minimization problem
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &= \frac{1}{2} \sum_{i=0}^2\abs{(u-\overline u)(x_i)}^2 + \frac{\alpha}{2}\|q\|^2&&\\
\text{s.t. } (\nabla u,\nabla \phi) &= (f(q),\phi) \quad\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with zero Dirichlet boundary conditions and
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\subsubsection{Program description}
{
In contrast to the first example, the control is now a discrete quantity (for
the three observation points) where we use the \texttt{FE\underline{}Nothing}
element to assign the three controls. Here, the number of components equals
the number of controls. 
In addition, }notice that the cost functional is of mixed type (from our computational point of view), i.e. the first part is a pointfunctional whereas the regularization part requires the evaluation of a domain integral. To handle this, we need a special integrator as well as a special newtonsolver. Additionally, our \texttt{LocalFunctional} returns as his type:
\begin{verbatim}
      string
      GetType() const
      {
        return "point domain";
      }
\end{verbatim}
This brings along that we have not only to implement 
\begin{itemize} 
\item \texttt{ElementValue},
\item \texttt{ElementValue\_U},
\item \texttt{ElementValue\_Q}, 
\item \texttt{ElementValue\_UU},
\item \texttt{ElementValue\_UQ},
\item \texttt{ElementValue\_QU},
\item \texttt{ElementValue\_QQ}
\end{itemize}
 from \texttt{FunctionalInterface}, but also all the aforementioned methods with a preceding \text{Point} (\texttt{PointValue} etc.).
%FE_Nothing, handle of the control erlaeutern.

>>>>

\subsubsection{General problem description}
This example solves the following pointwise minimization problem
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &= \frac{1}{2} \sum_{i=0}^2\abs{(u-\overline u)(x_i)}^2 + \frac{\alpha}{2}\|q\|^2&&\\
\text{s.t. } (\nabla u,\nabla \phi) &= (f(q),\phi) \quad\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with zero Dirichlet boundary conditions and
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\subsubsection{Program description}
{
In contrast to the first example, the control is now a discrete quantity (for
the three observation points) where we use the \texttt{FE\underline{}Nothing}
element to assign the three controls. Here, the number of components equals
the number of controls. 
In addition, }notice that the cost functional is of mixed type (from our computational point of view), i.e. the first part is a pointfunctional whereas the regularization part requires the evaluation of a domain integral. To handle this, we need a special integrator as well as a special newtonsolver. Additionally, our \texttt{LocalFunctional} returns as his type:
\begin{verbatim}
      string
      GetType() const
      {
        return "point domain";
      }
\end{verbatim}
This brings along that we have not only to implement {\color{red}the methods
\begin{align*} 
 &\texttt{ElementValue},&&
 \texttt{ElementValue\_U},\\
 &\texttt{ElementValue\_Q}, &&
 \texttt{ElementValue\_UU},\\
 &\texttt{ElementValue\_UQ},&&
 \texttt{ElementValue\_QU},\\
 &\texttt{ElementValue\_QQ},
\end{align*}}
 from \texttt{FunctionalInterface}, but also all the aforementioned methods with a preceding \text{Point} (\texttt{PointValue} etc.).
%FE_Nothing, handle of the control erlaeutern.
